|

### 2022-07-20 manuscript_BE/manuscript_BE.Rmd
New version of the manuscript

---
title: Field and experimental data together with computational 
  models reveal how cleaner fish adjust decisions  in a biological market
author:
  - name: Andrés E. Quiñones 
    email: andresqp@gmail.com
    affiliation: a
    corresponding: andresqp@gmail.com
  - name: Zegni Triki
    email: zegni.triki@gmail.com
    affiliation: b
  - name: Redouan Bshary
    email: redouan.bshary@unine.ch
    affiliation: Institute of Biology, University of Neuchatel
address:
  - code: a
    address: Institute of Biology, University of Neuchatel, Rue Emile-Argand 11, 2000-CH Neuchatel, Switzerland
  - code: b
    address: Department of Zoology, Stockholm University, Svante Arrheniusväg 18 B,  106 91 Stockholm, Sweden,
abstract: |
  While it is generally straightforward to quantify individual performance in cognitive experiments, identifying the underlying cognitive processes remains a major challenge. Often, different mechanistic underpinnings yield similar performances, and Lloyd Morgan’s cannon warrants acceptance of the simpler explanation. Alternatively, when the different mechanisms interact with environmental conditions, variation in performance across environments might allow to statistically infer the mechanism responsible. We illustrate this point by fitting computational models to experimental data on performance by wild-caught cleaner fish *Labroides dimidiatus* in an ephemeral reward task, as well as cleaner and client fish densities from the locations of capture. Using Bayesian statistics to fit the model parameters to performance data revealed that cleaner fish most likely estimate future consequences of an action, while it appears unlikely that the removal of the ephemeral reward acts as psychological punishment (negative reinforcement). Incorporating future consequences also yields performances that can be considered the result of locally optimal decision-rules, in contrast to the negative reinforcement mechanism. We argue that the combination of computational models with data is a powerful tool to infer the mechanistic underpinnings of cognitive performance.

author_summary: |
  Performance in behavioural experiments is often used to assess the cognitive abilities of animals. However, animals can get to the same outcome in alternative ways. Thus, the outcome of the experiments does not tell us how animals achieve their performance.  In order to overcome this limitation, we used a set of computational models, which provide predictions on how alternative cognitive mechanisms perform in the face of varying environmental conditions, and compare their predictions to performance data from individuals that experienced different conditions throughout their life. Our study system is the cleaner fish *Labroides dimidiatus*, a coral reef fish that feeds off ectoparasites and dead tissue from the skin of other reef fishes. Cleaner fish often have to choose among alternative clients seeking their cleaning service, and the optimal choice is often hard to achieve. Our performance experiments mimic the natural choices the cleaner fish make. The combination of experiments and computational models points to cleaner fish abilities to account for the future effect of their choices by estimating the long term expected value of those choices. Estimating long term value is a mechanism involved in human foresight.

bibliography: Cleanerlearning.bib
header-includes: |
  ```{=latex}
  \usepackage{float}
  \usepackage{booktabs}
  \newcommand{\beginsupplement}{ \setcounter{table}{0}     \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
  ```
output: rticles::plos_article
# plos.csl can be download and used locally
csl: http://www.zotero.org/styles/plos
---

_Text based on plos sample manuscript, see <https://journals.plos.org/ploscompbiol/s/latex>_

@@

Often alternative cognitive mechanisms yield similar behaviour and/or cognitive
performances. This poses a problem for disentangling the mechanistic 
underpinnings of behaviour. This is particularly clear in research aimed at
discovering between species variation in *higher* cognitive abilities; or 
in other words, research on whether non-human animals show cognitive 
abilities believed to be uniquely human. For instance, when researchers 
try to find *mental time travel-like* behaviour, they usually come up with 
clever experiments to show the behaviour displayed requires inferences made 
through past events[@dally_Foodcaching_2006]. However, they often face the
challenge of alternative scenarios where simpler explanations, 
like classic associative learning, can bring about the 
observed behavioural outcome [@suddendorf_Evolution_2007]. Similarly, attempts 
to demonstrate the presence of *theory of mind* in non-human animals 
face objections justified by alternative mechanisms underpinning 
similar behavioural results [@heyes_Theory_1998]. 
Such controversies are usually settled by using the principle 
of parsimony and its cognitive version, 
Lloyd Morgan’s cannon, which states that the simpler explanation (mechanism)
should be accepted. Ideally, however, alternative hypotheses should be evaluated 
in light of their explanatory power. One potential dimension that could allow
the variation necessary among distinct cognitive mechanisms is environmental
variation. Controlled laboratory conditions often mask between and within 
species variation that arises among different environments. Thus, if 
predictions from the alternative mechanisms depend on environmental 
conditions, it is possible to assess their relative explanatory power 
by performing cognitive test in the context of
different environmental conditions.

From an evolutionary perspective, mechanisms are likely selected because 
of how they allow individuals to respond to environmental variation. 
For example, biological market theory predicts that the exchange rate of goods 
and/or services traded between cooperative partners adjusts to the 
law of supply and demand, when individuals have some degree of partner choice
[@noe_Biological_1995]. Supply and demand conditions, which typically 
depend on the abundance of the species involved, certainly vary in time 
and space. Therefore, natural selection should favour the ability 
to flexibly adjust decisions and behavioural output to 
current market conditions. Indeed, such adjustments have been repeatedly 
documented [@axen_Signalling_1996]. In animals, an obvious general candidate 
mechanism for the strategic adjustment is the cognitive machinery. However, 
it is not clear which cognitive mechanisms allow individuals to adjust their 
behaviour to the varying conditions. More specifically the question arises 
to what extent mechanisms beyond basic associative learning may be involved.

One example of strategic adjustment in a biological market is the marine
cleaning mutualism involving the cleaner fish *Labroides dimidiatus* and
'client' fish. Client fish seek cleaner fish services at their territory 
(so-called “cleaning station”) and offer themselves as food patches 
to get their ectoparasites removed, which provides cleaners 
with food and clients with improved health [@waldie_LongTerm_2011;
@ros_Does_2011;@triki_Effects_2016;@demaire_Reduced_2020]. 
Given the capacity of some client fish to swim larger distances and 
access multiple cleaning stations while others access the only cleaning 
station in their territory, it is crucial to categorize clients as 
either “visitors” or “residents”, respectively. During cleaning interactions, 
a cleaner fish often face a choice between a visitor and a resident client 
seeking its cleaning services simultaneously. Visitors have the option 
to switch to another cleaner fish if being made to wait, while residents 
must wait for inspection. Indeed,
visitors have been observed to use their partner choice option in that
way [@bshary_Choosy_2002], which may explain why cleaners give
visitors service priority in a field study in the Red Sea
[@bshary_Cleaner_2001].

When researchers aimed at testing wild-caught cleaner fish, as 
well as individuals from other species, abilities 
to prioritize visitor over resident clients in a lab-based paradigm, 
they used Plexiglas plates of different colours and shapes offering the same 
amount of food as surrogates for visitor and resident clients.
One plate is made to behave like a resident, i.e. it remains until 
the cleaner fish has eaten off it. The other plate was made to behave 
like a visitor, i.e. it is removed if not inspected first. 
This laboratory paradigm originally inspired by the cleaning market 
of cleaner fish is more generally referred as the ephemeral reward task. 
Individuals are assessed in their capacity to develop a preference for the 
ephemeral option, which gives them access to two reward units.
Cleaners learned to prefer the visitor plate and
hence obtained the double amount of food [@bshary_Asymmetric_2002].
Furthermore, adult cleaner fish outperformed various primates as well as rats and
pigeons in this original version of the ephemeral reward task
[@salwiczek_Adult_2012; @zentall_Early_2017]; yet the African grey
parrots solve this task as well [@pepperberg_Can_2014]. 
This between variation in performance, and primates' known cognitive abilities
[@reader_Evolution_2001], shows that performance in the task
cannot be readily explained by classifying species as having high and 
low cognitive abilities. Instead, this between species variation suggests
that solving the ephemeral reward task requires a set of cognitive 
mechanisms that are not present in all the tested taxa, or were not used 
under the experimental set-up.

Recent research on the cognitive abilities needed to solve the ephemeral 
reward task has used a broad approach as proposed by Shettleworth 
[@shettleworth_Cognition_2009], who defined cognition as including 
all ways in which animals acquire information 
through the senses, process, retain and decide to act on it. 
Indeed, perception of relevant cues is of major importance 
for individual performance. For example, some animals find relevant 
information in the plates  [@wismer_Cuebased_2019] while others find 
it in the food [@pretot_Comparative_2021; @pretot_Comparing_2016]. 
However, identifying cues as salient is not the only challenge 
of the ephemeral reward task, as revealed by proximate learning models. 
Such models allow varying the cognitive tool kit and evaluating 
which minimal kit is necessary to solve 
the task at hand (e.g. [@dubois_Model_2021]). Applied to the ephemeral 
reward task, learning models showed that basic reinforcement learning does 
not suffice to solve the ephemeral reward task [@prat_Modelling_2022;
@quinones_Reinforcement_2019]. 
This is particularly so when models assume the more complex natural 
situation in which cleaner fish face not only resident-visitor pairs but also 
visitor-visitor and resident-resident pairs, as well a resident alone or 
a visitor alone. To be able to give visitors priority over residents, 
cleaners need to be able to assess a client’s value separately for the 
three possible scenarios (alone, paired with a fish with the 
same strategic option, paired with a fish with the alternative strategic option) 
[@quinones_Reinforcement_2019]. The ability to distinguish and value 
one stimulus differently alone from compound versions of it has been termed 
configurational learning, chunking, 
or segmentation (see references in [@prat_Modelling_2022]).

In addition to configurational learning, cleaners also need to account for 
the future consequences of current decisions. In the model by 
Quiñones *et al.* [@quinones_Reinforcement_2019], this could be achieved in 
two non-mutually exclusive ways: through low temporal discounting of 
future effects, also termed ‘chaining’ [@enquist_Power_2016]; and/or 
through perceiving a visitor client leaving as psychological punishment 
(i.e. as a negative reinforcer). Low temporal discounting is when 
individuals include in their valuation of an action the reward effects 
that this will have in the future. This is done by combining in a single 
valuation the reward obtained in the current time with all the reward that 
comes after, discounting for how far in the future reward 
is accrued. 'Chaining' the reward of these different time steps allows 
individuals to take actions that increase the long-term reward at the sacrifice
of short term considerations [@enquist_Power_2016]. Even though, 'chaining' 
can be readily implemented computationally in learning models 
[@enquist_Power_2016;@sutton_Reinforcement_2018], cognitively it seems to be 
a complex adaptation [@suddendorf_Evolution_2007]. On the other hand,
using client behaviour as a negative reinforcer is, in principle, 
easier to implement. Negative reinforcement is part of the ubiquitous 
learning mechanism termed operant conditioning [@thorndike_Animal_1898; 
@skinner_Behavior_1938]. Thus, the standard logic of 
Lloyd Morgan’s cannon demands that operant conditioning as the simpler 
explanation is to be accepted by default. Ideally, however, the two 
mechanisms should be evaluated in light of how well 
they explain the available data. Note that different fields interested in
cognition and decision making use different words to refer to negative 
reinforcers [@quinones_Reinforcement_2019;@sutton_Reinforcement_2018]. 
Here, for the sake of simplicity and clarity, we will use the word 
'penalty' to refer to this mechanism which includes a negative reinforcer.

Over the last decade, our team tested over a hundred wild-caught 
cleaner fish in the exact same paradigm of the ephemeral reward 
task [@salwiczek_Adult_2012;@wismer_Variation_2014;@triki_Decrease_2018;
@triki_Biological_2019;@triki_Brain_2020]. These fish often come from 
different reef locations around the study field site at Lizard Island, 
Great Barrier Reef in Australia. Further investigation of the local 
eco-sociological conditions revealed that cleaner and client 
fish population densities have a substantial impact on cleaner fish 
performance in the task. Cleaner fish from reef sites with relatively 
low densities were more likely to fail at solving the task [@triki_Biological_2019;
@triki_Decrease_2018; @wismer_Variation_2014]. The explanation for this pattern
is twofold. First, low cleaner fish population densities imply 
low supply and high demand for cleaning services. 
Under such conditions, there are fewer occasions under 
which failing to give priority to a visitor translates to an empty 
cleaning station. Second and also related to the state of the cleaning market,
visitor clients at low cleaner fish population densities 
are less likely to exert their partner choice options 
and are hence more likely to wait for inspection if not given cleaning 
priority [@triki_Brain_2020]. In the absence of partner choice 
cleaner fish should not give priority to visitors any more.

The interplay between the underlying cognitive machinery and 
local ecological conditions apparently generate the documented variable 
performance among individuals of the same species. 
Our approach of fitting the computational model to the empirical data on 
fish densities and cleaner fish censuses and performance
in the ephemeral reward task aimed at: (i), determining which 
mechanism cleaner fish use to incorporate future consequences of current 
decisions by testing whether chaining, penalty, or a combination 
of both best explains their performance; (ii)  determining whether the 
two mechanisms differed with respect to the ecological conditions 
that are likely to cause high versus low performance in the 
ephemeral reward task. Additionally, we assessed which mechanism 
yields optimal performance patterns. Relying on the logic of 
biological market theory, we predicted 
that appropriate performance is to show a low preference for visitors under low
local cleaner-to-client ratio.

@@

The model consists of a set of individual-based simulations where
individuals, representing cleaner fish, face a series of choices 
between two options, which simulate the natural conditions 
of the cleaning market. Individuals experience a series 
of discrete time points in which they face different ‘states’, 
defined by the number and category of client fish (visitor or residents) 
inviting for cleaning services. There are six possible states: 
zero clients, one resident, one visitor, 
resident-resident, visitor-visitor, and resident-visitor. 
The probability of each state is largely determined by the relative 
abundance of cleaner fish, residents and visitors, but to some degree by
cleaner fish choices when it faces the resident-visitor combination. 
This is because residents are willing to queue for cleaning service; 
while visitors leave the queue (with a certain probability) when made to wait. 
Individuals obtain a fixed reward from cleaning a client fish
regardless of the category. Every time individuals face and make a choice
they update the probability of making that same choice. The update is
based on the difference between the expected reward and the obtained
reward (prediction error) [@sutton_Reinforcement_2018; 
@rescorla_Theory_1972]. Given the new information, the update is carried
in the direction that leads to more reward being obtained, given
the new information. In the long run, the probability of choosing a visitor 
over a resident converges in the model. To which probability the model 
will converge depends on the relative abundance of cleaners, visitors 
and residents; as well as on the probability of visitors leaving the 
cleaning station when made to wait. Further details of the model 
implementation can be found in Quiñones *et al* [@quinones_Reinforcement_2019].

The model shows that agents need to find a way 
to incorporate future consequences of current choices. In the model, 
this could be achieved with either of two parameters that could 
also work together.  First, $\gamma$ measures how
much individuals include future rewards in their decision updates. 
If $\gamma=0$, individuals only use the immediate reward obtained from a
cleaning interaction. As $\gamma$ increases, individuals include more the
reward obtained from the subsequent choices. That amounts to calculating
and using for decision making the future expected rewards of an action (chaining).
Second, $\eta$ measures how much individuals include in their reward the
fleeing behaviour of visitors as a negative component (penalty). Both of
these parameters allow individuals to use in their estimates the future
effects of their choices.

@@

The empirical data were collected between 2010 and 2019 always during 
the austral winter months June to August from a total of five 
study reef sites (Corner Beach-CB, Horseshoe-HS, Mermaid Cove-MC, 
Northern Horseshoe-NHS, and The Crest-TC)  at Lizard Island 
($14.6682^\circ S, 145.4604^\circ E$), Great Barrier Reef, Australia. 
The data consist of three sets: fish censuses, field 
observations of cleaner-client interactions to quantify the probability 
of visitors leaving if made to wait, and the 
performance of wild-caught cleaner fish in the ephemeral reward test. 
In total, we have twelve site/year data sets for fish censuses and 
corresponding performance in the lab test. Thus, some sites were sampled more 
than once. To estimate the population density of cleaner fish 
and their clients at a given site in 
a given year, Triki *et al.* [@triki_Biological_2019] used a series of ten 
transects of $30m$ each. Observers swam along the transect lines placed on 
the reef and first counted the visible large-bodied adult fish 
(species with total length $TL \geq 10cm$) including cleaner fish 
on a width of $5m$, and then on the 
return individuals of small-bodied fish species ($TL < 10 cm$) \
on a width of $1 m$
(see Triki et al. 2018 for further details on fish censuses data collection). 
We then scaled the counts of cleaner fish, small-bodied, 
and  large-bodied  clients fish densities per $100 m^2$. 
From the study by Triki *et al* [@triki_Biological_2019] 
cleaner fish and large-bodied client populations densities were 
highly correlated, and only the former 
was hence used in the analyses as representative of both measures.

The field observation data consisted of video recordings/encodings of the 
cleaner-client cleaning interactions. There were videos from eight 
cleaners per site/year of a duration of $30 min$ each. Triki *et al* 
extracted information from every event wherein a visitor client 
was made to wait in favour of another client (visitor or resident), and noted 
whether or not the visitor left or queued for the cleaning service
[@triki_Biological_2019;@triki_Brain_2020].

>>>>

---
title: | 
  \large \textbf{Title:} Computational models reveal how cleaner fish adjust decisions in a biological market
subtitle: |
  \normalsize \textbf{Abbreviated title:} Cleaner fish cognitive mechanisms in a biological market
# date: "`r format(Sys.time(), '%B %d, %Y')`"
# author: 
#   - name: Andrés E. Quiñones 
#     email: andresqp@gmail.com
#     affiliation: a
#     corresponding: andresqp@gmail.com 
#     #, Rue Emile-Argand 11, CH-2000, Neuchatel, Switzerland. +41 32 718 31 24
#   - name: Zegni Triki
#     affiliation: a
#     affiliation: b
#   - name: Redouan Bshary
#     affiliation: a
# address:
#   - code: a
#     address: Institute of Biology, University of Neuchatel, Neuchatel, Switzerland
#   - code: b
#     address: Department of Zoology, Stockholm University, Stockholm, Sweden
output:
   pdf_document:
    keep_tex: true
    latex_engine: xelatex
    template: template.latex
fig_caption: yes
fontsize: 12pt
keep_tex: yes
bibliography: Cleanerlearning.bib
style: M:\Zotero\styles\behavioral-ecology.csl
spacing: double
abstract: |
  While it is generally straightforward to quantify individual performance in cognitive experiments, identifying the underlying cognitive processes remains a major challenge. Often, different mechanistic underpinnings yield similar performances, and Lloyd Morgan’s cannon warrants acceptance of the simpler explanation. Alternatively, when the different mechanisms interact with environmental conditions, variation in performance across environments might allow to statistically infer the mechanism responsible. We illustrate this point by fitting computational models to experimental data on performance by wild-caught cleaner fish *Labroides dimidiatus* in an ephemeral reward task, as well as cleaner and client fish densities from the locations of capture. Using Bayesian statistics to fit the model parameters to performance data revealed that cleaner fish most likely estimate future consequences of an action, while it appears unlikely that the removal of the ephemeral reward acts as psychological punishment (negative reinforcement). Incorporating future consequences also yields performances that can be considered the result of locally optimal decision-rules, in contrast to the negative reinforcement mechanism. We argue that the combination of computational models with data is a powerful tool to infer the mechanistic underpinnings of cognitive performance. 
lay_summary: |
  Cleaner fish eat ectoparasites off other fishes, so-called clients. It regularly happens that two clients seek a cleaner's service simulatenously. Cleaners benefit from prioritising clients unwilling to wait, so they can feed on those willing to wait. To make the right choice, cleaners must somehow “look” into the future to anticipate consequences of current choices. By combining a learning model with data, we show that cleaners estimate the long-term value of their actions rather than using simpler heuristics. Estimating long-term value is a mechanism involved in human foresight.
header-includes: |
  ```{=latex}
  \usepackage{float}
  \usepackage{graphicx}
  \usepackage{booktabs}
  \linespread{2}
  \usepackage{lineno}
  \linenumbers
  \usepackage{setspace}\doublespacing
  \newcommand{\beginsupplement}{ \setcounter{table}{0}     \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
  \providecommand{\keywords}[1]{\textbf{\textit{Keywords---}} #1}
  ```
editor_options:
  markdown:
    wrap: 80
---

```{=latex}
\keywords{learning, behavior, cleaners, bayesian statisitics, behavioral mechanisms}
```

@@

Often alternative cognitive mechanisms yield similar behavior and/or cognitive
performances. This poses a problem for disentangling the mechanistic 
underpinnings of behavior. This is particularly clear in research aimed at
discovering between species variation in *higher* cognitive abilities; or 
in other words, research on whether non-human animals show cognitive 
abilities believed to be uniquely human. For instance, when researchers 
try to find *mental time travel-like* behavior, they usually come up with 
experiments to show the behavior displayed requires inferences made 
through past events [@dally_Foodcaching_2006]. However, they often face the
challenge of alternative scenarios where simpler explanations, 
like classic associative learning, can bring about the 
observed behavioral outcome [@suddendorf_Evolution_2007]. Similarly, attempts 
to demonstrate the presence of *theory of mind* in non-human animals 
face objections justified by alternative mechanisms underpinning 
similar behavioral results [@heyes_Theory_1998]. 
Such controversies are usually settled by using the principle 
of parsimony and its cognitive version, 
Lloyd Morgan’s cannon, which states that the simpler explanation (mechanism)
should be accepted. Ideally, alternative hypotheses should be evaluated 
in light of their explanatory power.

<!-- If alternative mechanisms predict different  -->
<!-- outcomes when exposed to different contexts, the variation arising from -->
<!-- varying the context could open up a way to infer which mechanism is used.  -->
<!-- Controlled laboratory conditions often mask between and within  -->
<!-- species variation that arises among different environmental contexts. Thus, if  -->
<!-- predictions from the alternative mechanisms depend on environmental  -->
<!-- conditions, it is possible to assess their relative explanatory power  -->
<!-- by performing cognitive tests in the context of different  -->
<!-- environmental conditions.  -->

Learning is a key overarching cognitive mechanism that allows 
individuals to associate rewards with environmental stimuli 
and thus behave adaptively [@staddon_Adaptive_2016;@shettleworth_Cognition_2009]. 
Associative learning, in particular, exists in all major vertebrate taxa, 
and in many invertebrates as well [@heyes_Simple_2012;@macphail_Brain_1982; 
@staddon_Adaptive_2016; @behrens_Associative_2008]. 
Associative learning is not homogeneous throughout its taxonomic distribution, 
rather there are differences across and within species 
[@shettleworth_Cognition_2009;@sih_Linking_2012]. So presumably, 
the mechanistic underpinnings of learning have been modified by 
natural selection [@marler_Species_1989].

One way to formalize the alternative mechanistic underpinnings of 
associative learning is to develop quantitative models of learning 
processes. This approach, which started within experimental psychology 
[@staddon_Adaptive_2016;@brandon_Computational_2002], has been very fruitful 
in disentangling the mechanistic structure of cognitive systems. 
More recently, the development of reinforcement learning theory 
[@sutton_Reinforcement_2018], has allowed to 
evaluate the empirical support for alternative mechanistic hypotheses by
providing quantitative predictions which are amenable to statistical tests
[@farashahi_Learning_2020;@farashahi_Featurebased_2017]. Interestingly, these 
learning models not only have received support from behavioral data, but also 
are consistent with the current view on reward processing in the brain 
[@schultz_Neuronal_2015].

From an evolutionary perspective, mechanisms are likely selected because 
of how they allow individuals to respond to environmental variation. 
For example, biological market theory predicts that the exchange rate of goods 
and/or services traded between cooperative partners adjusts to the 
law of supply and demand, when individuals have some degree of partner choice
[@noe_Biological_1995a]. Supply and demand conditions, which typically 
depend on the abundance of the species involved, certainly vary in time 
and space. Therefore, natural selection should favor the ability 
to flexibly adjust decisions and behavioral output to 
current market conditions. Indeed, such adjustments have been 
documented [@axen_Signalling_1996]. One example of strategic 
adjustment in a biological market is the marine
cleaning mutualism involving the cleaner fish *Labroides dimidiatus* and
'client' fishes. Client fishes seek cleaner fish services at their territory 
(so-called “cleaning station”) and offer themselves as food patches 
to get their ectoparasites removed, which provides cleaners 
with food and clients with improved health [@waldie_LongTerm_2011;
@ros_Does_2011;@triki_Effects_2016;@demaire_Reduced_2020]. 
Given the capacity of some client fish to swim larger distances and 
access multiple cleaning stations while others access the only cleaning 
station in their territory, it is crucial to categorize clients as 
either “visitors” or “residents”, respectively. During cleaning interactions, 
a cleaner fish often faces a choice between a visitor and a resident client 
seeking its cleaning services simultaneously. Visitors have the option 
to switch to another cleaner fish if being made to wait, while residents 
must wait for inspection. Indeed,
visitors have been observed to use their partner choice option in that
way [@bshary_Choosy_2002], which may explain why cleaners give
visitors service priority in a field study in the Red Sea
[@bshary_Cleaner_2001a]. Furthermore, in a lab based paradigm, design 
to mimic the resident-visitor choice (ephemeral reward task), cleaners learned 
to prefer the cue associated with the epheral food source (visitor) and
hence accessed both food sources, obtaining 
double the amount of food [@bshary_Asymmetric_2002].
However, further exploration revealed that not all cleaner fish 
manage to develop a preference for the ephemeral option 
in the lab [@triki_Decrease_2018]. Over the last decade, over a hundred 
wild-caught cleaner fish have been tested
in the exact same paradigm of the ephemeral reward 
task [@salwiczek_Adult_2012;@wismer_Variation_2014;@triki_Decrease_2018;
@triki_Biological_2019;@triki_Brain_2020]. These fish often come from 
different reef locations. Investigation of the local 
eco-sociological conditions revealed that cleaner and client 
fish population densities have a substantial impact on cleaner fish 
performance in the task. Cleaner fish from reef sites with relatively 
low densities were more likely to fail at solving the task 
[@triki_Biological_2019;
@triki_Decrease_2018; @wismer_Variation_2014]. This intra-specific variation
is unlikely due to local genetic adaptation, because cleaner
fish are open water spawners and the environmental conditions can vary within 
the lifespan of a fish.

Mechanistic models explicitly designed to mimic the ephemeral reward task 
have shown that the simplest form of associative learning 
(operant conditioning) cannot account for a solution to 
the ephemeral reward task [@prat_Modelling_2022;
@quinones_Reinforcement_2019]. Operant conditioning is 
a form of associative learning where individuals use short term 
reward to associate and choose actions. Such models allow varying 
the cognitive tool kit and evaluating which minimal kit is necessary 
to solve the task at hand (e.g. @dubois_Model_2021). 
To be able to give visitors priority over residents, 
cleaners need to be able to assess a client’s value separately for the 
three possible scenarios (alone, paired with a fish with the 
same strategic option, paired with a fish with the alternative strategic option) 
[@quinones_Reinforcement_2019]. The ability to distinguish and value 
one stimulus differently alone from compound versions of it has been termed 
configurational learning, chunking, or 
segmentation (see references in @prat_Modelling_2022). 
In addition to configurational learning, cleaners also need to account for 
the future consequences of current decisions. In the model by 
@quinones_Reinforcement_2019, this could be achieved in 
two non-mutually exclusive ways: through low temporal discounting of 
future effects, also termed ‘chaining’ [@enquist_Power_2016]; and/or 
through perceiving a visitor client leaving as psychological punishment 
(i.e. as a negative reinforcer). Chaining is when 
individuals include in their valuation of an action the reward effects 
that this will have in the future. This is done by combining in a single 
valuation the reward obtained in the current time with all the reward that 
comes after, discounting for how far in the future reward 
is accrued. 'Chaining' the reward of these different time steps allows 
individuals to take actions that increase the long-term reward at the sacrifice
of short term considerations [@enquist_Power_2016]. Even though, 'chaining' 
can be readily implemented computationally in learning models 
[@enquist_Power_2016;@sutton_Reinforcement_2018], cognitively it seems to be 
a complex adaptation [@suddendorf_Evolution_2007]. On the other hand,
using client behavior as a negative reinforcer is, in principle, 
easier to implement. Thus, the standard logic of 
Lloyd Morgan’s cannon demands that operant conditioning as the simpler 
explanation is to be accepted by default. Ideally, however, the two 
mechanisms should be evaluated in light of how well 
they explain the available data. Note that different fields interested in
cognition and decision making use different words to refer to negative 
reinforcers [@quinones_Reinforcement_2019;@sutton_Reinforcement_2018]. 
Here, for the sake of simplicity and clarity, we will use the word 
'penalty' to refer to this mechanism which includes a negative reinforcer.

In here we used the field and experimental data to fit the parameters of 
a reinforcement learning model to infer  the 
cognitive mechanism that cleaners use in their interaction with clients. 
Specifically, our approach of fitting the computational model 
to the empirical data aimed at: (i), determining which 
mechanism cleaner fish use to incorporate future consequences of current 
decisions by testing whether chaining, penalty, or a combination 
of both best explains their performance; (ii)  determining whether the 
two mechanisms differed with respect to the ecological conditions 
that are likely to cause high versus low performance in the 
ephemeral reward task. Additionally, we assessed which mechanism 
yields optimal performance patterns. Relying on the logic of 
biological market theory, we predicted that appropriate performance 
is to show a high preference for visitors only under high
local cleaner-to-client ratio.

@@

The model consists of a set of individual-based simulations where
individuals, representing cleaner fish, face a series of choices 
between two options, which simulate the natural conditions 
of the cleaning market. Individuals experience a series 
of discrete time points in which they face different ‘states’, 
defined by the number and category of client fish (visitor or residents) 
inviting for cleaning services. There are six possible states: 
zero clients, one resident, one visitor, 
resident-resident, visitor-visitor, and resident-visitor. 
The probability of each state is largely determined by the relative 
abundance of cleaner fish, residents and visitors, but to some degree by
cleaner fish choices when it faces the resident-visitor combination. 
This is because residents are willing to queue for cleaning service; 
while visitors leave the queue (with a certain probability) when made to wait. 
Individuals obtain a fixed reward from cleaning a client fish
regardless of the category. Every time individuals face and make a choice
they update the probability of making that same choice. The update is
based on the difference between the expected value and the obtained
reward - the prediction error ($\delta_t$) - [@sutton_Reinforcement_2018; 
@rescorla_Theory_1972]. Formally, the prediction error is given by

\begin{equation}
\label{eq:delta}
\delta_t = R_t -  V_t(S_t) + \gamma V_t(S_{t+1}) ,
\end{equation}

where $R_t$ is the sum different reward sources at time $t$; 
$V_t(S_t)$ is the estimated value at time $t$ of the the state faced by the agent 
at time $t$; similarly $V_t(S_{t+1})$ is the estimated value of the state
to come in the following time-step, $\gamma$ is the discount factor for
future rewards. When the estimated value of the current state ($V_t(S_t)$) 
is equal to the sum of short-term ($R_t$) and future discounted reward 
($\gamma V_t(S_{t+1})$) learning stops for that state. If $\gamma = 0$ 
the estimates made by the agent only capture short-term reward. We assume 
short-term reward to two components: positive reward determined by the 
amount of food obtained from cleaning a client; and negative reward triggered
when by a client leaving the station without being cleaned. Formally, 
we let total reward be given by $R_t=P_t - \eta_t$. Where $\eta$ is a parameter
of the model that determines the the size of the negative reward triggered by
unattended clients leaving the station.

The prediction error (Eq. \ref{eq:delta}) is used to update the value of each one
of the states the agent faces, as well as the preference for the 
resident and visitor options. The value update is simple the product of the 
prediction error and the parameter for the speed of learning 
($\Delta V(S_t)=\alpha \delta_t$). The change in preference between the resident 
and visitor is given by $\Delta (\theta_v-\theta_r)_t=\alpha\delta_t2(1-\pi_v)$,
where $\theta_i$ represent the preference for one of two options and the 
difference captures the total change relative to one another; 
$\pi_v$ corresponds to the current probability of choosing the visitor. 
$p_i$ is determined by applying the logistic function to the difference in
preferences between the two mutually exclusive options 
($\pi_v=\frac{1}{1+e^{-(\theta_v-\theta_r)}}$). This amounts
to a preference update that is carried in the direction that 
leads to more reward being obtained, given
the new information. In the long run, the probability of choosing a visitor 
over a resident converges in the model. To which probability the model 
will converge depends on the relative abundance of cleaners, visitors 
and residents; as well as on the probability of visitors leaving the 
cleaning station when made to wait. Further details of the model 
implementation can be found in @quinones_Reinforcement_2019.

The model shows that agents need to find a way 
to incorporate future consequences of current choices. In the model, 
this could be achieved with either of two parameters that could 
also work together.  First, $\gamma$ measures how
much individuals include future rewards in their decision updates. 
If $\gamma=0$, individuals only use the immediate reward obtained from a
cleaning interaction. As $\gamma$ increases, individuals include more the
reward obtained from the subsequent choices. That amounts to estimating
and using for decision making the future expected rewards of an action (chaining).
Second, $\eta$ measures how much individuals include in their reward the
fleeing behavior of visitors as a negative component (penalty). Both of
these parameters allow individuals to use in their estimates the future
effects of their choices.

@@

The empirical data were collected between 2010 and 2019 always during 
the austral winter months June to August from a total of five 
study reef sites (Corner Beach-CB, Horseshoe-HS, Mermaid Cove-MC, 
Northern Horseshoe-NHS, and The Crest-TC)  at Lizard Island 
($14.6682^\circ S, 145.4604^\circ E$), Great Barrier Reef, Australia. 
The data consist of three sets: fish censuses, field 
observations of cleaner-client interactions to quantify the probability 
of visitors leaving if made to wait, and the 
performance of wild-caught cleaner fish in the ephemeral reward test. 
In total, we have twelve site/year data sets for fish censuses and 
corresponding performance in the lab test. Thus, some sites were sampled more 
than once. To estimate the population density of cleaner fish 
and their clients at a given site in 
a given year, @triki_Biological_2019 used a series of ten 
transects of $30m$ each. Observers swam along the transect lines placed on 
the reef and first counted the visible large-bodied adult fish 
(species with total length $TL \geq 10cm$) including cleaner fish 
on a width of $5m$, and then on the 
return individuals of small-bodied fish species ($TL < 10 cm$)
on a width of $1 m$
(see @triki_Decrease_2018 for further details on fish censuses data collection). 
We then scaled the counts of cleaner fish, small-bodied, 
and  large-bodied  clients fish densities per $100 m^2$. 
From the study by @triki_Biological_2019 
cleaner fish and large-bodied client populations densities were 
highly correlated, and only the former 
was hence used in the analyses as representative of both measures.

The field observation data consisted of video recordings/encodings of the 
cleaner-client cleaning interactions. There were videos from eight 
cleaners per site/year of a duration of $30 min$ each. Triki et al. 
extracted information from every event wherein a visitor client 
was made to wait in favour of another client (visitor or resident), and noted 
whether or not the visitor left or queued for the cleaning service
[@triki_Biological_2019;@triki_Brain_2020].
